Estimating duration using survival analysis

### 2023-08-11 cis-perfect-testing.tex
Use (un)detected rather than (un)truncated

The first complication in the analysis is that infection episodes can be undetected, meaning no there are no positive tests associated with them.
This happens when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
For these undetected episodes, there is no information on them episode, meaning that we do not even know how many of them occurred.
It is important to account for undetected episodes because they are systematically of shorter duration than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration was 24 days (or longer), the individual would still have been detectable at their next test, and therefore would have been detected.
Undetected infections are a large problem in CIS because the interval between tests is often 28 days (see\todo{ref section on the survey design}), large relative to the duration of an infection episode.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/truncation}
  \caption[Truncation in CIS data]{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

@@

In this section, I develop a model accounting for double interval censoring and truncation, modelling only individuals with detected episodes.
This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}; in particular, I adopt the framework of \textcite{heiseyModelling}.
The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
Previous studies consider the case where nothing was observed regarding individuals without a positive test (left truncation).
I do not make use of this additional information here because it makes little difference but increases the computational requirements and strength of assumptions regarding reinfections/immunity required (see \cref{E-appendix:total-model} for details).
The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were undetected.

@@

\begin{itemize}
\item
  Admissible episodes, $\alpha_i$, which have an infection and
  recovery time consistent with the data observed. That is, $b_i \in [l_i^{(b)}, r_i^{(b)}]$, and $e_i \in [l_i^{(e)}, r_i^{(e)}]$.
  $n_{ia} =1$ of these occurred, each with probability
  $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
\item
  Truncated episodes, $\Omega_i^C$, which have an infection and
  recovery time such that they would not have tested positive. An
  unknown number, $n_{it}$ of these observed, each with probability
  $p_{it} = \prob((b, e) \in \Omega^C_i \mid \theta)$.
\item
  Inadmissible episodes, $\beta_i$, which are not consistent with the data observed but would have been
  observed (not truncated). This is all remaining episodes not in the
  previous sets. $n_{iu} = 0$ of these occurred, where they would have
  occurred with probability
  $p_{iu} = \prob((b, e) \in \beta_i \mid \theta)$.
\end{itemize}

The untruncated region (where we could have observed an episode) is
$\Omega_i = \alpha_i \cup \beta_i$ (the notation here is used for
consistency with \textcite{heiseyModelling}), and is the
complement of $\Omega^C_i$.

\begin{figure}
\includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
\caption[Admissible, inadmissible, and untruncated infections]{The regions (as defined in the main text) for an episode $i$ in an
individual whose first test was at time 0 and negative, with subsequent
negative tests at times 7, 14, and 56, and positive tests at times 21
and 28. Inadmissible region is unshaded. The black (impossible) region is because it does not make sense for recovery to occur prior to infection. \label{perf-test:fig:partitionSpace}}
\end{figure}

@@

Next, we derive $p_{it} = \prob((b, e) \in \Omega^C_i)$, the probability
that episode $i$ was truncated. A truncated episode means that no test
was performed during the episode or there was no negative test prior to
the episode. Denote by $t_i$ the set of testing times for the
individual in which episode $i$ was observed, and define the time
until the next test on the individual after time $t'$ as:
\begin{align}
t^N_{it} &= \min \{ t' \in t_i : t' \geq t \} - t.
\intertext{Then:}
\Omega^C_i
&= \{ (b, e) : \nexists t \in t_i. b \leq t \leq e \vee b \leq \min(t_i) \} \\
&= \{ (b, e) : e - b < t^N_{ib} \vee b \leq \min(t_i) \}.
\intertext{Hence:}
1 - p_{it}
&= 1 - \prob(E_i - B_i < t_{iB_i}^N \vee B_i \leq \min(t_i)) \\
&= 1 - \prob(E_i - B_i < t_{iB_i}^N \wedge B_i > \min(t_i)) - \prob(B_i \leq \min(t_i)) \\
&= 1 - \sum_{b=\min(t_i) + 1}^T \prob(E_i - b + 1 < t_{ib}^N \mid B_i = b) \prob(B_i = b) \\
  &\; - \sum_{t=1}^{\min(t_i)} \prob(B_i = b)\\
&= 1 - \frac{1}{T} \sum_{b=\min(t_i)+1}^T (1 - S_\theta(t_{ib}^N + 1)) - \frac{\min(t_i)}{T} \\
&= 1 - \frac{T-\min(t_i)}{T} + \frac{1}{T} \sum_{b=\min(t_i)+1}^T S_\theta(t_{ib}^N + 1)) - \frac{\min(t_i)}{T} \\
&= \frac{1}{T} \sum_{b=\min(t_i)+1}^T S_\theta(t_{ib}^N + 1)
\end{align}

@@

To generate the durations, an assumption is required for the distribution of the duration of positivity.
I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to represent what is seen within CIS.
The tail is modified based on an unpublished analysis of CIS data by Sarah Walker (described below).
Sarah uses survival analysis to estimate the duration from CIS data.
The initiating event is assumed known as the time the episode was detected.
The final event is assumed interval-censored between the time of the final positive test and the subsequent negative test, or right-censored if a negative test has not yet been observed.
A flexible, spline-based form is assumed for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
By not accounting for either the truncation or the interval censoring of the initiating event, this analysis has competing biases which makes them hard to interpret~\autocite{cisMethodsONS}.

>>>>

The first complication in the analysis is that infection episodes can be undetected, meaning no there are no positive tests associated with them.
This happens when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
For these undetected episodes, there is no information on them episode, meaning that we do not even know how many of them occurred.
It is important to account for undetected episodes because they are systematically of shorter duration than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration was 24 days (or longer), the individual would still have been detectable at their next test, and therefore would have been detected.
Undetected infections are a large problem in CIS because the interval between tests is often 28 days (see\todo{ref section on the survey design}), large relative to the duration of an infection episode.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/truncation}
  \caption[Undetected episodes in CIS data]{An unknown number of infection episodes in CIS are undetected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

@@

In this section, I develop a model accounting for double interval censoring and undetected episodes, modelling only individuals with detected episodes.
This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}; in particular, I adopt the framework of \textcite{heiseyModelling}.
The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
Previous studies consider the case where nothing was observed regarding individuals without a positive test (left truncation).
I do not make use of this additional information here because it makes little difference but increases the computational requirements and strength of assumptions regarding reinfections/immunity required (see \cref{E-appendix:total-model} for details).
The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were undetected.

@@

\begin{itemize}
\item
  Admissible episodes, $\alpha_i$, which have an infection and
  recovery time consistent with the data observed. That is, $b_i \in [l_i^{(b)}, r_i^{(b)}]$, and $e_i \in [l_i^{(e)}, r_i^{(e)}]$.
  $n_{ia} =1$ of these occurred, each with probability
  $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
\item
  Undetected (or, for consistency with the literature, truncated) episodes, $\Omega_i^C$, which have an infection and
  recovery time such that they would not have tested positive. An
  unknown number, $n_{it}$ of these observed, each with probability
  $p_{it} = \prob((b, e) \in \Omega^C_i \mid \theta)$.
\item
  Inadmissible episodes, $\beta_i$, which are not consistent with the data observed but would have been
  detected. This is all remaining episodes not in the
  previous sets. $n_{iu} = 0$ of these occurred, where they would have
  occurred with probability
  $p_{iu} = \prob((b, e) \in \beta_i \mid \theta)$.
\end{itemize}

The detected region (where we could have observed an episode) is
$\Omega_i = \alpha_i \cup \beta_i$ (the notation here is used for
consistency with \textcite{heiseyModelling}), and is the
complement of $\Omega^C_i$.

\begin{figure}
\includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
\caption[Admissible, inadmissible, and undetected infections]{The regions (as defined in the main text) for an episode $i$ in an
individual whose first test was at time 0 and negative, with subsequent
negative tests at times 7, 14, and 56, and positive tests at times 21
and 28. Inadmissible region is unshaded. The black (impossible) region is because it does not make sense for recovery to occur prior to infection. \label{perf-test:fig:partitionSpace}}
\end{figure}

@@

Next, we derive $p_{it} = \prob((b, e) \in \Omega^C_i)$, the probability
that episode $i$ was detected. A detected episode means that no test
was performed during the episode or there was no negative test prior to
the episode. Denote by $t_i$ the set of testing times for the
individual in which episode $i$ was observed, and define the time
until the next test on the individual after time $t'$ as:
\begin{align}
t^N_{it} &= \min \{ t' \in t_i : t' \geq t \} - t.
\intertext{Then:}
\Omega^C_i
&= \{ (b, e) : \nexists t \in t_i. b \leq t \leq e \vee b \leq \min(t_i) \} \\
&= \{ (b, e) : e - b < t^N_{ib} \vee b \leq \min(t_i) \}.
\intertext{Hence:}
1 - p_{it}
&= 1 - \prob(E_i - B_i < t_{iB_i}^N \vee B_i \leq \min(t_i)) \\
&= 1 - \prob(E_i - B_i < t_{iB_i}^N \wedge B_i > \min(t_i)) - \prob(B_i \leq \min(t_i)) \\
&= 1 - \sum_{b=\min(t_i) + 1}^T \prob(E_i - b + 1 < t_{ib}^N \mid B_i = b) \prob(B_i = b) \\
  &\; - \sum_{t=1}^{\min(t_i)} \prob(B_i = b)\\
&= 1 - \frac{1}{T} \sum_{b=\min(t_i)+1}^T (1 - S_\theta(t_{ib}^N + 1)) - \frac{\min(t_i)}{T} \\
&= 1 - \frac{T-\min(t_i)}{T} + \frac{1}{T} \sum_{b=\min(t_i)+1}^T S_\theta(t_{ib}^N + 1)) - \frac{\min(t_i)}{T} \\
&= \frac{1}{T} \sum_{b=\min(t_i)+1}^T S_\theta(t_{ib}^N + 1)
\end{align}

@@

To generate the durations, an assumption is required for the distribution of the duration of positivity.
I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to represent what is seen within CIS.
The tail is modified based on an unpublished analysis of CIS data by Sarah Walker (described below).
Sarah uses survival analysis to estimate the duration from CIS data.
The initiating event is assumed known as the time the episode was detected.
The final event is assumed interval-censored between the time of the final positive test and the subsequent negative test, or right-censored if a negative test has not yet been observed.
A flexible, spline-based form is assumed for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
By not accounting for either the undetected infections or the interval censoring of the initiating event, this analysis has competing biases which makes them hard to interpret~\autocite{cisMethodsONS}.
